
\documentclass[11pt]{article}

\setlength{\oddsidemargin}{0in}
\setlength{\evensidemargin}{0in}
\setlength{\topmargin}{-0.3in}
\setlength{\headheight}{0in}
\setlength{\headsep}{0in}
\setlength{\textwidth}{6.5in}
\setlength{\textheight}{9.3in}
\setlength{\parindent}{0in}
\setlength{\parskip}{0.1in}

\usepackage{times}
\usepackage{hyperref}

\begin{document}

% \begin{rawhtml}
% <body bgcolor=white>
% 
% \end{rawhtml}

\title{Norman S. Matloff}

\author{Professor of Computer Science \\
University of California at Davis} 

\date{}

\maketitle

\section*{Education } 

Ph.D. in (Pure) Mathematics, UCLA, 1975

% \section*{Positions Held} 
% 
% % \begin{itemize}
% 
% % \item 
% {\it University of California, Davis}:  
% Faculty member 1975-present, starting in Dept. of Mathematics, later
% Dept. of Statistics, then Dept. of Computer Science. 
% 
% % \item 
% {\it Remote Computing Corporation, Palo Alto, CA}: 
% Senior Software Project Leader, 1979-1980.  
% Directed a large project in the development of a national
% financial securities database system.  
% 
% % \end{itemize}

\section*{Current Research Interests} 

Data mining, machine learning, random graph modeling, parallel
processing software and hardware, simulation.

\section*{Recent Publications} 

{\it The Art of Debugging, with GDB, DDD and Eclipse} (book), No Starch
Press, with Peter Salzman, 2008.

Virtual Migration, by A. Aneesh, invited book review by N. Matloff, {\it
Journal of International Migration and Integration}, December 2008,
425-427.

A New Method for Rule Finding Via Bootstrapped Confidence Intervals.
{\it SIAM Conference on Data Mining}, 2008, 547-552.

Using Soft-line Recursive Response to Improve Query Aggregation in
Wireless Sensor Networks (with X. Lu, M. Spear, K. Levitt, F. Wu).  {\it
Proceedings of the 2008 IEEE International Conference on
Communications}, 2309-2316.

Synchronization Attack and Defense in Energy-Efficient Listen-Sleep
Slotted MAC Protocols (with X. Lu, M. Spear, K. Levitt, F. Wu),  {\it
SECURWARE Archive Proceedings of the 2008 Second International
Conference on Emerging Security Information, Systems and Technologies},
403-411.

Availability-Aware Provisioning Strategies for Differentiated Protection
Services in Wavelength-Convertible WDM Mesh Networks (with B. Mukherjee,
H. Zang, J. Zhang and K. Zhu). {\it IEEE/ACM Transactions on Networking},
2007, 15, 5, 1177-1190.

Revisiting the Issue of Performance Enhancement of Discrete Event
Simulation Software (with Alex Bahouth, Steven Crites, and Todd
Williamson), {\it Proceedings of the 40th Anuual Simulation Symposium,}
2007, 114-122.

{\it From Algorithms to Z-Scores: Probabilistic and Statistical Modeling
in Computer Science} (open textbook), 2006,
\url{http://heather.cs.ucdavis.edu/probstatbook}.

On the Adverse Impact of Work Visa Programs on Older U.S. Engineers and
Programmers (invited paper).  {\it California Labor and Employment Law
Review} (a publication of the California State Bar Association), August
2006.

\section*{Recent Grants}

{\it Evaluation of and Enhancement to the SimPy Simulation Language for
Use in DNT, NHI, and E\&E Applications}, PI N. Matloff, \$17,000, 2006.

{\it Davis Social Links}, PI F. Wu, Co-PIs N. Matloff and M. Neff, FIND,
\$700,000, 2008.

\section*{Campus Awards}

\begin{itemize}

\item Campus-wide Distinguished Teaching Award, 2006

\item 
Campus-wide Distinguished Public Service Award, 2002

\item 
Campus-wide Outstanding Faculty Advisor Award, 1994

\end{itemize}

\end{document}


